Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic

### 2024-03-26 paper.Rmd
typos, labels in proofread

Theoretically, @liebigSociologyJustice2016 establishes three fundamental distinctions in justice attitudes in the field of sociological justice research. First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement. Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities). Finally, they can be _outcome-related attitudes_ related primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes). Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]

@@

In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are ISSP (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the ESS 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects. The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.

@@

The perception of economic inequality is one of the factors associated with the justification of educational inequality. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the undersetimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]

@@

The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

@@

Finally, and in line with hypothesis 1 and 4, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged. As a consequence, the justification of inequality could show a decrease over time:

@@

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The waves 2016,2017,2018,2019 and 2022 were administered using computer-assisted personal interviewing (CAPI). However, in 2021 a reduced version was conducted using computer-assisted telephone interviews (CATI). In addition, wave 3 included a refreshment sample to counter survey attrition. The same sampling strategy as that of wave 1 was implemented for selecting new cases. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends. Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants. In broader terms, the accumulated attrition between Wave 1 and Wave 6 is 41%, achieving a final sample of N = 1728. The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.

@@

The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills"). In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large"). All these variables were also measured using a Likert scale from "strongly disagree" (1) to "strongly agree" (5). Then,  educational level and household income quintiles are included as objective status measures, as well as subjective social status. Finally, political identification on a left-right scale, age and gender were included as control variables. Table \@ref(tab:descriptives) shows the items used, their response categories, and their frequency.

@@

Given the hierarchical structure of the data (observations nested in survey waves), we applied a longitudinal multilevel strategy [@singer_applied_2009]. Longitudinal multilevel models are suited to account for the shared variance among units in the data to estimate standard errors better. Since individuals share variance within themselves over time, if the error structure is not considered, it would be as if they were considered different individuals. Multilevel models allow a solution in regression estimation by adding a random term that represents the variance associated with the nesting of the data (random effects). The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].

@@

Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022. Each year represents stacked percentual frequencies, and the flows in between represent the within-subjects’ change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010]. For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories. In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave. This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.

@@

Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves. In this matrix, the correlations vary between low and moderate values. Justification of economic inequality depicts a moderate and negative association with the perception of inequality and a moderate and positive association with both meritocracy perception variables. Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables. Finally, the two meritocracy variables present a high and positive association with each other.

@@

Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<0.01). However, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<0.01), but it has a larger weight than the perception of talent.

@@

omit <- "(education)|(ess)"
texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
                omit.coef = omit,
                custom.coef.names = c("Intercept",
                                      "Wave 2017",
                                      "Wave 2018",
                                      "Wave 2019",
                                      "Wave 2022",
                                      "Wave",
                                      "Wave^2",
#                                      "High School <br> <i>(Ref. Primary)</i>",
#                                      "Technical",
#                                      "Universitary",
                                      "Quintile 2",
                                      "Quintile 3",
                                      "Quintile 4",
                                      "Quintile 5",
                                      "Quintile NA",
#                                      "Subjective Social Status",
                                      "Age 30-49",
                                      "Age 50-64",
                                      "Age 65 or more",
                                      "Gender (Ref. Male)",
                                      "Center",
                                      "Right",
                                      "Does not identify",
                                      "Inequality perception",
                                      "Merit: Talent",
                                      "Merit: Effort" ),
                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
                  "Quintile (Ref.= Quintile 1)" = 8:12,
                  "Age (Ref.= 18-29)" = 13:15,
                  "Pol. pos (Ref.= Left)" = 17:19
                  ),
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual"),
               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.765,
               float.pos = "!ht", 
               booktabs = FALSE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE)

@@

In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time. In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic. We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact). Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for). Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort. As observed, only the effort variable shows a significant interaction with the time variable (0.01, p<0.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.

@@

Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic. In general, our results showed mixed evidence about our hypotheses. Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal. We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens. In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic. As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises. As in the Chilean case the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies. A closer look to the political environment during the last year in the country could give additional hints. Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly. This could have driven preferences in a conservative direction, which was actually reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].

@@

Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time. However, longitudinal evidence shows that the perception of meritocracy linked effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$). In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era. An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort-, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers. Again, this certainly needs further studies that include additional specific conceptual and methodological approach to this phenomenon.

Among the limitations of our study, we can mention at least three. First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies. In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions. Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital. In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard. Finally, there are limitations involving the longitudinal dimension of the data used. Given limitations for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately the educational justice indicator was excluded in that survey. Consequently, the results about temporal change have to be considered carefully, and further data waves could give us more information in this matter.

The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates. In this regard, thanks to the survey's sampling strategy, it would be possible to make contextual estimates at the municipality level in future studies. For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.  Furthermore, this dataset could  be used for comparing inequality justification in policy areas besides education, such as in health and pensions. Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.

@@

# (APPENDIX) Appendix {.unnumbered}

@@

omit <- "(education)|(ess)"
texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
#                omit.coef = omit,
                custom.coef.names = c("Intercept",
                                      "Wave 2017",
                                      "Wave 2018",
                                      "Wave 2019",
                                      "Wave 2022",
                                      "Wave",
                                      "Wave^2",
                                      "High School",
                                      "Technical",
                                      "Universitary",
                                      "Quintile 2",
                                      "Quintile 3",
                                      "Quintile 4",
                                      "Quintile 5",
                                      "Quintile NA",
                                      "Subjective Social Status",
                                      "Age 30-49",
                                      "Age 50-64",
                                      "Age 65 or more",
                                      "Gender (Ref. Male)",
                                      "Center",
                                      "Right",
                                      "Does not identify",
                                      "Inequality perception",
                                      "Merit: Talent",
                                      "Merit: Effort" ),
                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
                              "Education (Ref.=Primary)"= 8:10,
                  "Quintile (Ref.= Quintile 1)" = 11:15,
                  "Age (Ref.= 18-29)" = 17:19,
                  "Pol. pos (Ref.= Left)" = 21:23
                  ),
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual"),
               caption = paste("(\\#tab:multilevel-full)","Multilevel effects of justification of economic inequality"),
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.73,
               float.pos = "!ht", 
               booktabs = FALSE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE)

>>>>

 Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research. First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement. Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities). Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes). Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]

@@

In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects. The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.

@@

The perception of economic inequality is one of the factors associated with the justification of educational inequality. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]

@@

The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

@@

Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged. As a consequence, the justification of inequality could show a decrease over time:

@@

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI). However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021. In addition, wave 3 included a refreshment sample to counter survey attrition. The same sampling strategy as that of wave 1 was implemented for selecting new cases. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends. Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants. In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728. The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.

@@

The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills"). In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large"). All these variables were also measured using a Likert scale from "strongly disagree" (1) to "strongly agree" (5). Then,  educational level and household income quintiles are included as objective status measures, as well as subjective social status. Finally, political identification on a left-right scale, age, and gender were included as control variables. Table \@ref(tab:descriptives) shows the items used, their response categories, and their frequency.

@@

Given the hierarchical structure of the data (observations nested in survey waves), we applied a longitudinal multilevel strategy [@singer_applied_2009]. In this matter, longitudinal multilevel models are suited to account for the shared variance among units in the data to estimate standard errors better as observations are nested within respondents. Since individuals share variance within themselves over time, if the error structure is not considered, it would be as if they were considered different individuals. Multilevel models allow a solution in regression estimation by adding a random term that represents the variance associated with the nested structure of the data (random effects). Consequently, the linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].

@@

Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022. Each year represents stacked percentual frequencies, and the flows in between represent the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010]. For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories. In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave. This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.

@@

Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves. In this matrix, the correlations vary between low and moderate values. Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01). Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01). Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).

@@

Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01). However, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.

@@

omit <- "(education)|(ess)"
texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
                omit.coef = omit,
                custom.coef.names = c("Intercept",
                                      "Wave 2017",
                                      "Wave 2018",
                                      "Wave 2019",
                                      "Wave 2022",
                                      "Wave",
                                      "Wave^2",
#                                      "High School <br> <i>(Ref. Primary)</i>",
#                                      "Technical",
#                                      "Universitary",
                                      "Quintile 2",
                                      "Quintile 3",
                                      "Quintile 4",
                                      "Quintile 5",
                                      "Quintile (No information)",
#                                      "Subjective Social Status",
                                      "Age 30-49",
                                      "Age 50-64",
                                      "Age 65 or more",
                                      "Gender (Ref. Male)",
                                      "Center",
                                      "Right",
                                      "Does not identify",
                                      "Inequality perception",
                                      "Merit: Talent",
                                      "Merit: Effort" ),
                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
                  "Household Income (Ref.= Quintile 1)" = 8:12,
                  "Age (Ref.= 18-29)" = 13:15,
                  "Pol. pos (Ref.= Left)" = 17:19
                  ),
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual"),
               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.765,
               float.pos = "!ht", 
               booktabs = FALSE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE)

@@

In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time. In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic. We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact). Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for). Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort. As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.

@@

Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic. In general, our results showed mixed evidence about our hypotheses. Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal. We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens. In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic. As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises. As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies. A closer look to the political environment during the last year in the country could give additional hints. Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly. This could have driven preferences in a conservative direction, which was actually reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].

@@

Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time. However, longitudinal evidence shows that the perception of meritocracy linked effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$). In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era. An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers. Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.

Among the limitations of our study, we can mention at least three. First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies. In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions. Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital. In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard. Finally, there are limitations involving the longitudinal dimension of the data used. Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey. Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.

The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates. In this regard, thanks to the sampling strategy the survey, it would be possible to make contextual estimates at the municipality level in future studies. For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.  Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions. Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.

@@

\newpage 
# (APPENDIX) Appendix {.unnumbered}

@@

omit <- "(education)|(ess)"
texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
#                omit.coef = omit,
                custom.coef.names = c("Intercept",
                                      "Wave 2017",
                                      "Wave 2018",
                                      "Wave 2019",
                                      "Wave 2022",
                                      "Wave",
                                      "Wave^2",
                                      "High School",
                                      "Technical",
                                      "Universitary",
                                      "Quintile 2",
                                      "Quintile 3",
                                      "Quintile 4",
                                      "Quintile 5",
                                      "Quintile (No information)",
                                      "Subjective Social Status",
                                      "Age 30-49",
                                      "Age 50-64",
                                      "Age 65 or more",
                                      "Gender (Ref. Male)",
                                      "Center",
                                      "Right",
                                      "Does not identify",
                                      "Inequality perception",
                                      "Merit: Talent",
                                      "Merit: Effort" ),
                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
                              "Education (Ref.=Primary)"= 8:10,
                  "Household Income (Ref.= Quintile 1)" = 11:15,
                  "Age (Ref.= 18-29)" = 17:19,
                  "Pol. pos (Ref.= Left)" = 21:23
                  ),
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual"),
               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.60,
               float.pos = "!ht", 
               booktabs = FALSE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE)
